\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-03 introduction.tex
tooling
exhaustivity

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and instance variables from its superclass.
But the language-theoretic essence of inheritance is more general:
it is a linguistic approach to \emph{incrementally modifying
recursive definitions}~\cite{cook1990inheritance}.

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97}, % and associated types~\cite{ckj05},
extensible cases~\cite{bac2006}, etc.\ 
are all based on this essential idea of inheritance.
%
In particular, when a language mechanism allows inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

@@

Integrating family polymorphism into a dependent type theory for
logical reasoning, however, poses significant technical challenges.
As we analyze, pillars of dependent type theories—including
inductive types, definitional equality, and logical consistency—are
all inimical to the kind of extensibility and family polymorphism
found in existing OO language designs.
Thus, our contributions include novel design recipes for dealing with
these challenges and strong meta\-theoretical guarantees on the
underlying logic.

@@

Specifically, we make the following contributions.

>>>>

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and instance variables from its superclass.
But the language-theoretic essence of inheritance is more general:
it is a linguistic approach to incrementally modifying
record-like definitions by allowing \emph{late binding}~\cite{cook1990inheritance}.

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97}, % and associated types~\cite{ckj05},
extensible cases~\cite{bac2006}, etc.\ 
are all based on this essential idea of inheritance.
%
In particular, when a language mechanism allows late binding of the
meaning of nested types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

@@

Integrating family polymorphism into a dependent type theory for
logical reasoning, however, poses significant technical challenges.
As we analyze, pillars of dependent type theories—including
inductive types, definitional equality, and logical consistency—are
all inimical to the kind of extensibility and family polymorphism
found in existing OO language designs.
Thus, our contributions include novel design recipes for dealing with
these challenges and strong meta\-theoretical guarantees on the
underlying logic.
%
Specifically, we make the following contributions.

@@

### 2022-10-03 lang-design.tex
exhaustivity

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLC_fix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types ("FInductive") and
adds new cases to the recursive functions ("FRecursion") and induction proofs ("FInduction").
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.

\noindentparagraph{Inductive types are extensible.}

In family "SLTC_fix", inductive type "tm" is formed by the four
constructors inherited from "STLC" and a fifth constructor "tm_fix".

Crucially, the types of the constructors are polymorphic to their enclosing family.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lstinline[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lstinline{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "SLTC_fix" include those constructed from "tm_fix".

\noindentparagraph{Induction remains exhaustive.}

\ref{chg:extensible-v-exhaustive}

>>>>

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLC_fix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.

\noindentparagraph{Extensibility of inductive types.}

In family "STLC_fix", inductive type "tm" \emph{further binds} the "tm" type
in family "STLC". It is generated by five constructors: four inherited from
"STLC" and a fifth called "tm_fix".

Crucially, the types of the constructors are polymorphic to their enclosing family---%
their meaning is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lstinline[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLC_fix" include those constructed from "tm_fix".

\noindentparagraph{Exhaustivity of induction.}

Ordinarily, an inductive type is not extensible: it is exhaustively
generated by its constructors and has no more inhabitants beyond those
they construct.
This idea is captured by the eliminator (aka recursor) associated with
an inductive type.
For example, if "tm" were defined as an ordinary inductive type, then
its eliminator would have the following type:

\begingroup
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall (P : tm -> Type), [@\hlcase{tm-recursor-case-var}@](\forall x, P (tm_var x))[@\tikzmarkend{tm-recursor-case-var}@] -> [@\hlcase{tm-recursor-case-unit}@]P tm_unit[@\tikzmarkend{tm-recursor-case-unit}@] -> [@\hlcase{tm-recursor-case-abs}@](\forall x t, P t -> P (tm_abs x t))[@\tikzmarkend{tm-recursor-case-abs}@] ->
  [@\hlcase{tm-recursor-case-app}@](\forall t1 t2, P t1 -> P t2 -> P (tm_app t1 t2))[@\tikzmarkend{tm-recursor-case-app}@] -> \forall t, P t
\end{lstlisting}
\endgroup

\noindent
The eliminator would allow for function definitions by recursion
and proofs by induction, over "tm", that exhaustively
handle the four cases corresponding to each constructor.
The (dependent) return type~"P" is called the \emph{motive} of the recursion.
%Exhaustivity of induction would follow, trivially.

With inductive types made extensible, exhaustivity of induction becomes
in question, however.
To reconcile the tension without requiring redefinition or rechecking of
case handlers (challenge \ref{chg:extensible-v-exhaustive}),
our design introduces the \lsti{FRecursion} and \lsti{FInduction} commands.
The key idea is to allow case handlers to be added retroactively, should
inductive types be extended, and to allow recursion and induction (which are defined in terms of case handlers) to be late bound.

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC_fix" further binds that in "STLC":
the four cases from "STLC" are automatically inherited and reused,
with "STLC_fix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

Like every other field of the family, the meaning of "subst" is late bound.
For example, the type of "st_beta" (modeling the $\beta$-reduction rule)
in family "STLC" refers to "subst".
When "st_beta" is inherited into family "STLC_fix", it has a type that now
refers to the "subst" function in the same family, where "subst" is defined
by handling all the five cases known to that family.

The \lstinline{FInduction} command is similar to \lstinline{FRecursion} but
allows cases to be defined in proof mode.
For example, "weakenlem" is proven by induction on the typing relation
"hasty".
Its \lsti{motive} clause suggests that the lemma reads as
\lsti{\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T}.
Upon entering case "ht_abs", for instance, Coq enters proof mode with the goal

\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
  \forall ...
\end{lstlisting}

\noindent
The programmer can discharge the goal using tactic programming.
The idiomatic programming experience with Coq is largely preserved.

### 2022-10-03 stlc-mechanized.tex
exhaustivity

FInductive tm : Set :=[@\hfill\codecomment{Terms}@]
| tm_var  : id -> tm
| tm_unit : tm
| tm_abs  : id -> tm -> tm
| tm_app  : tm -> tm -> tm.[@\label{line:tm-app}@]

@@

FInductive hasty : env -> tm -> ty -> Prop :=
| ht_var  : [@\dadada@]  | ht_unit : [@\dadada@][@\hfill\codecomment{Typing rules}@]
| ht_abs  : [@\dadada@]  | ht_app  : [@\dadada@].

FInductive value : tm -> Prop :=[@\hfill\codecomment{Value forms}@]
| v_unit : value tm_unit
| v_abs  : \forall x e, value (tm_abs x e).

FRecursion subst on tm[@\hfill\codecomment{Substitution function}@] 
motive \lambda(_ : tm), id -> tm -> tm.
Case tm_var := \lambda y x e,
  if (eqb x y) then e else (tm_var y).
Case tm_unit := \lambda x e, tm_unit.
Case tm_abs := \lambda y e' IHe' x e, [@\dadada@].
Case tm_app := \lambda e1 IHe1 e2 IHe2 x e,
  tm_app (IHe1 x e) (IHe2 x e).
End subst.

FInductive step : tm -> tm -> Prop :=
| st_app1 : [@\dadada@][@\hfill\codecomment{Reduction rules}@]
| st_app2 : [@\dadada@]
| st_beta : \forall x e v, value v ->
  step (tm_app (tm_abs x e) v) (subst e x v).

@@

FInduction weakening on hasty[@\hfill\codecomment{Weaken.\ lemma}@]
motive \lambda G e T (_ : hasty G e T),
  \forall G', includedin G G' -> hasty G' e T.
Case ht_var.  [@\dadada@] Qed.[@\hfill@]Case ht_unit. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app.  [@\dadada@] Qed.
End weakening.

FInduction substlemma on hasty[@\hfill\codecomment{Subst.\ lemma}@]
motive \lambda G' e T (_ : hasty G' e T),
  \forall G x e' T', G' = extend G x T' ->
  hasty empty e' T' -> hasty G (subst e x e') T.
Case ht_var.  [@\dadada@] Qed.[@\hfill@]Case ht_unit. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app.  [@\dadada@] Qed.
End subst_lemma.

FInduction preserve on hasty[@\hfill\codecomment{Preserv.\ theorem}@]
motive \lambda G e T (_ : hasty G e T),
  G = empty -> \forall e', step e e' -> hasty empty e' T.
Case ht_var.  [@\dadada@] Qed.[@\hfill@]Case ht_unit. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app.  [@\dadada@] Qed.
End preserve.

FInduction progress on hasty[@\hfill\codecomment{Progress theorem}@]
motive \lambda G e T (_ : hasty G e T),
  G = empty -> value e \/ \exists e', step e e'.
Case ht_var.  [@\dadada@] Qed.[@\hfill@]Case ht_unit. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app.  [@\dadada@] Qed.
End progress.

Field typesafe :[@\hfill\codecomment{Type-safety theorem}@]
  \forall e e' T, steps e e' -> hasty empty e T ->
  value e' \/ \exists e'', step e' e''.
Proof. [@\dadada@] Qed.

@@

FInductive hasty : env -> tm -> ty -> Prop +=
| ht_fix : \forall G x e T, hasty (extend G x T) e T ->
  hasty G (tm_fix x e) T.

FRecursion subst on tm motive \lambda _, id -> tm -> tm.
Case tm_fix := [@\dadada@].
End subst.

FInductive step : tm -> tm -> Prop +=
| st_fix : \forall x e,
  step (tm_fix x e) (subst e x (tm_fix x e)).

FInduction weakening on hasty motive [@\dadada@].
Case ht_fix. [@\dadada@] Qed.
End weakening.

FInduction substlemma on hasty motive [@\dadada@].
Case ht_fix. [@\dadada@] Qed.
End subst_lemma.

>>>>

FInductive tm : Set :=[@\hfill\codecomment{Terms}@]
| tm_unit : tm
| tm_var  : id -> tm
| tm_abs  : id -> tm -> tm
| tm_app  : tm -> tm -> tm.[@\label{line:tm-app}@]

@@

FInductive hasty : env -> tm -> ty -> Prop :=
| ht_unit : [@\dadada@]  | ht_var : [@\dadada@][@\hfill\codecomment{Typing rules}@]
| ht_abs  : [@\dadada@]  | ht_app : [@\dadada@].

FInductive value : tm -> Prop :=[@\hfill\codecomment{Value forms}@]
| v_unit : value tm_unit
| v_abs  : \forall x t, value (tm_abs x t).

FRecursion subst on tm[@\hfill\codecomment{Substitution function}@] 
motive \lambda(_ : tm), id -> tm -> tm.
Case tm_unit := \lambda x t, tm_unit.
Case tm_var := \lambda y x t,
  if (eqb x y) then t else (tm_var y).
Case tm_abs := \lambda y t' IHt' x t,
  tm_abs y (if (eqb x y) then t' else IHt' x t).
Case tm_app := \lambda t1 IHt1 t2 IHt2 x t,
  tm_app (IHt1 x t) (IHt2 x t).
End subst.

FInductive step : tm -> tm -> Prop :=
| st_app1 : [@\dadada@][@\hfill\codecomment{Reduction rules}@]
| st_app2 : [@\dadada@]
| st_beta : \forall x t v, value v ->
  step (tm_app (tm_abs x t) v) (subst t x v).

@@

FInduction weakenlem on hasty[@\hfill\codecomment{Weaken.\ lemma}@]
motive \lambda G t T (_ : hasty G t T),
  \forall G', includedin G G' -> hasty G' t T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End weakenlem.

FInduction substlem on hasty[@\hfill\codecomment{Subst.\ lemma}@]
motive \lambda G' t T (_ : hasty G' t T),
  \forall G x t' T', G' = extend G x T' ->
  hasty empty t' T' -> hasty G (subst t x t') T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End substlem.

FInduction preserve on hasty[@\hfill\codecomment{Preserv.\ theorem}@]
motive \lambda G t T (_ : hasty G t T),
  G = empty -> \forall t', step t t' -> hasty empty t' T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End preserve.

FInduction progress on hasty[@\hfill\codecomment{Progress theorem}@]
motive \lambda G t T (_ : hasty G t T),
  G = empty -> value t \/ \exists t', step t t'.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End progress.

Field typesafe :[@\hfill\codecomment{Type-safety theorem}@]
  \forall t t' T, steps t t' -> hasty empty t T ->
  value t' \/ \exists t'', step t' t''.
Proof. [@\dadada@] Qed.

@@

FInductive hasty : env -> tm -> ty -> Prop +=
| ht_fix : \forall G x t T, hasty (extend G x T) t T ->
  hasty G (tm_fix x t) T.

FRecursion subst on tm motive \lambda _, id -> tm -> tm.
Case tm_fix := \lambda y t' IHt' x t, [@\dadada@].
End subst.

FInductive step : tm -> tm -> Prop +=
| st_fix : \forall x t,
  step (tm_fix x t) (subst t x (tm_fix x t)).

FInduction weakenlem on hasty motive [@\dadada@].
Case ht_fix. [@\dadada@] Qed.
End weakenlem.

FInduction substlem on hasty motive [@\dadada@].
Case ht_fix. [@\dadada@] Qed.
End substlem.
